ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-27 papers/GPUScA-2011/intro.tex
spelling and ack
Final corrections from Craig on plane to Munich before submission.

The peak performance of these newer accelerator architectures can be substantial.
Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
processor while the performance of the M2090 NVIDIA Tesla processor is in the
same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
many of the new accelerator architectures offer comes at a cost.  Architectural
changes are trending toward multiple heterogeneous cores and less of a reliance
on superscalar instruction level parallelism and hardware managed memory
hierarchies (such as traditional caches).  These changes place a heavy burden on
application programmers as they are attempt to adapt to these new systems.  An
especially challenging problem is not only how to program to these new
architectures --- considering the massive scale of concurrency available --- but
also how to design programs that are portable across the changing landscape of
computer architectures.  How does a programmer write one program that can
perform well on both a conventional multicore CPU \emph{and} a GPU (or another
of the emerging many-core architectures)?

A directive-based approach, such as OpenMP or the Accelerator programming model
from the Portland Group~\cite{pgi10accelerator}, is one solution to this
problem.  However, in this paper we take a somewhat different approach.  A
common theme amongst the new processors is the emphasis on data-parallel
programming.  This model is well suited to architectures that are
based on either vector processing or massively parallel collections of simple
cores.  The recent CUDA and OpenCL programming languages are are intended to
support this programming model.

>>>>

The peak performance of these newer accelerator architectures can be substantial.
Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
processor while the performance of the M2090 NVIDIA Tesla processor is in the
same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
many of the new accelerator architectures offer comes at a cost.  Architectural
changes are trending toward multiple heterogeneous cores and less of a reliance
on superscalar instruction level parallelism and hardware managed memory
hierarchies (such as traditional caches).  These changes place a heavy burden on
application programmers as they work to adapt to these new systems.  An
especially challenging problem is not only how to program to these new
architectures --- considering the massive scale of concurrency available --- but
also how to design programs that are portable across the changing landscape of
computer architectures.  How does a programmer write one program that can
perform well on both a conventional multicore CPU \emph{and} a GPU (or another
of the emerging many-core architectures)?

A directive-based approach, such as OpenMP or the Accelerator programming model
from the Portland Group~\cite{pgi10accelerator}, is one solution to this
problem.  However, in this paper we take a somewhat different approach.  A
common theme amongst the new processors is the emphasis on data-parallel
programming.  This model is well suited to architectures that are
based on either vector processing or massively parallel collections of simple
cores.  The recent CUDA and OpenCL programming languages are intended to
support this programming model.

### 2011-06-27 papers/GPUScA-2011/programming-model.tex
Final corrections from Craig on plane to Munich before submission.

A question that one may pose is ``\emph{Why choose Fortran and not a more modern
  language like X for programming accelerator architectures?}''  The recent rise
in interest in concurrency and parallelism at the language level driven by
multicore CPUs and many-core accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.  However,
for many scientific users with existing codes written in Fortran, new languages
and language extensions to use novel new architectures present a challenge: how
do they effectively use them while avoiding rewriting code and potentially
growing dependent on a transient technology that will vanish tomorrow?
In this paper we explore the constructs in Fortran that are particularly
relevant to GPU architectures.

@@

Borrowing ideas from ZPL, we introduce a concept of a region to Fortran with a
set of functions that allow programmers to work with subarrays in expressions.
In Fortran, these functions return a copy of or a pointer to an existing array or array section.
This is unlike ZPL, where regions are analogous to index sets and are used
primarily for address resolution within an array without dictating storage
related behavior.  The functions that we propose are similar in that they allow
a programmer to deal with index regions that are meaningful to their algorithm,
and automatically induce a halo (or ghost) cell pattern as needed in the
implementation generated by the compiler, where the size of an array is
implicitly increased to provide extra array elements surrounding the interior 
portion of the array.
%%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.

Region functions are similar to the shift operator as they can be used to
reference portions of the array that are shifted with respect to the interior
portion.  However, unlike the shift operator, regions are not expressed in terms
of boundary conditions and thus don't explicitly \emph{require} a knowledge of,
nor the application of, boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which access
only local subsections of an array stored in global memory.

ForOpenCL provides two new functions that are defined in Fortran and are used
in array-syntax operations.  Each function takes a integer array halo argument
that specifies the number of ghost cells on either side of a region, for each
dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
for a two-dimensional region.  These functions are:

@@

\begin{itemize}
\item There are no loops or index variables to keep track of.  Off by
  one index errors and improper handling of array boundaries are a
  common programming mistake.
\item The written code is closer to the algorithm, easier to
  understand, and is usually substantially shorter.
\item Semantically the intrinsic functions return arrays by value.
  This is usually what the algorithm requires.
\item Pure and elemental function are free from side effects, so it is
  easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

>>>>

A question that one may pose is ``\emph{Why choose Fortran and not a more modern
  language like X for programming accelerator architectures?}''  The recent rise
in interest in concurrency and parallelism at the language level driven by
multicore CPUs and many-core accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.  However,
for many scientific users with existing codes written in Fortran, new languages
and language extensions to use novel new architectures present a challenge: how
do programmers effectively use them while avoiding rewriting code and potentially
growing dependent on a transient technology that will vanish tomorrow?
In this paper we explore the constructs in Fortran that are particularly
relevant to GPU architectures.

@@

Borrowing ideas from ZPL, we introduce a concept of a region to Fortran with a
set of functions that allow programmers to work with subarrays in expressions.
In Fortran, these functions return a copy of or a pointer to an existing array or array section.
This is unlike ZPL, where regions are analogous to index sets and are used
primarily for address resolution within an array without dictating storage
related behavior.  The functions that we propose are similar in that they allow
a programmer to deal with index regions that are meaningful to their algorithm,
and automatically induce a halo (or ghost) cell pattern as needed in the
implementation generated by the compiler, where the size of an array is
implicitly increased to provide extra array elements surrounding the interior 
portion of the array.  It is important to note, however, that all memory allocated
by the programmer must explicitly contain the extra array elements in the halo.
%%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.

Region functions are similar to the shift operator as they can be used to
reference portions of the array that are shifted with respect to the interior
portion.  However, unlike the shift operator, regions are not expressed in terms
of boundary conditions and thus don't explicitly \emph{require} a knowledge of,
nor the application of, boundary conditions locally (global boundary conditions must
be explicitly provided by the programmer outside of calls to kernel procedures).
Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which access
only local subsections of an array stored in global memory.

ForOpenCL provides two new functions that are defined in Fortran and are used
in array-syntax operations.  Each function takes an integer array halo argument
that specifies the number of ghost cells on either side of a region, for each
dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
for a two-dimensional region.  These functions are:

@@

\begin{itemize}
\item There are no loops or index variables to keep track of.  Off by
  one index errors and improper handling of array boundaries are a
  common programming mistake.
\item The written code is closer to the algorithm, easier to
  understand, and is usually substantially shorter.
\item Semantically the intrinsic function {\tt region\_cpy} returns an array by value.
  This is usually what the algorithm requires.
\item Pure and elemental functions are free from side effects, so it is
  easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

### 2011-06-27 papers/GPUScA-2011/source-to-source.tex
lots of edits
Final corrections from Craig on plane to Munich before submission.

OpenCL~\cite{opencl08} is an open-language standard for developing applications
targeted for GPUs, as well as for multi-threaded applications targeted for
multi-core CPUs.  The kernels are run by calling a C runtime library from the
OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
underway and Fortran interfaces to the C runtime (distributed in the ForOpenCL
library).

@@

We point out that the hierarchical distribution of memory used on the OpenCL
device shown in Figure~\ref{fig:cl-memory} is similar to the distribution of
memory across MPI nodes in an MPI application.  In the case of MPI, the
virtual global array is represented by the background layer (with its halo) and
partitions of the global array are stored in the 4 MPI nodes shown in the foreground.

@@

@@

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, the {\tt wave\_advance} interface declaration is transformed as

{\small
\begin{verbatim}
__kernel void
wave_advance(float dx, ..., __global float * H, ...);
\end{verbatim}
}

The intent(in) arrays have local equivalents that are stored in local memory
and are declared by, for example,
{\small
\begin{verbatim}
__local float H_local[LOCAL_SIZE];
\end{verbatim}
}
\noindent
These local arrays are declared with the appropriate size and are
copied to local memory by the compiler with an inlined library function.  The
array temporaries defined on cell faces are declared similarly while interior
pointer variables are simple scalars, e.g., {\tt float pH}.  Intent(in) array
variables cannot be scalar objects because regions may be shifted and thus
\emph{shared} by threads within a thread group.

>>>>

OpenCL~\cite{opencl08} is an open-language standard for developing applications
targeted for GPUs, as well as for multi-threaded applications targeted for
multi-core CPUs.  The kernels are run by calling a C runtime library from the
OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
underway and Fortran interfaces to the C runtime are distributed in the ForOpenCL
library.

@@

We point out that the hierarchical distribution of memory used on the OpenCL
device shown in Figure~\ref{fig:cl-memory} is similar to the distribution of
memory across MPI nodes in an MPI application.  In the case of MPI, the
virtual global array is represented by the background layer (with its halo) and
partitions of the global array are stored in the 4 MPI nodes shown in the foreground.
Our current and future work on this effort includes source-to-source transformations
to generate MPI code in addition to OpenCL in order to deal with clusters of 
nodes containing accelerators.  This work is outside the scope of this paper.

@@

\vskip .4in

@@

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, the {\tt wave\_advance} interface declaration is transformed as
{\small
\begin{verbatim}
__kernel void
wave_advance(float dx, ..., __global float * H, ...);
\end{verbatim}
}
\noindent
The intent(in) arrays have local equivalents that are stored in local memory
and are declared by, for example,
{\small
\begin{verbatim}
__local float H_local[LOCAL_SIZE];
\end{verbatim}
}
\noindent
These local arrays are declared with the appropriate size and are
copied to local memory by the compiler with an inlined library function.  The
array temporaries defined on cell faces are declared similarly while interior
pointer variables are simple scalars, e.g., {\tt float pH}.  Intent(in) array
variables cannot be scalar objects because regions may be shifted and thus
\emph{shared} by threads within a thread group.
